---
editor_options:
  markdown:
    wrap: 72
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r,include=FALSE}
knitr::opts_chunk$set(digits.signif = 4, tidy = TRUE, tidy.opts=list(blank = FALSE, warning = FALSE), message = TRUE, echo = TRUE, width='80%', widt.cutoff = '80')
```

# S&P 500 Top 50 Components by Market Cap

```{r,include=FALSE}
library(janitor)
library(gridExtra)
library(grid)
library(broom)
library(cluster)
library(tidyverse)

library(plotly)
library(ggplot2)

library(factoextra)
library(xts)
library(TTR)
library(quantmod)
library(dplyr)
library(yfR)
library(rvest)
library(knitr)
library(tidyquant)
library(purrr)
library(writexl)
library(tidyr)
library(knitr)
library(dplyr)
library(timeR)
conflicted::conflict_prefer("mutate", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")
conflicted::conflict_prefer("lag", "dplyr")
library(rcartocolor)
nColor <- 12
ticker_colors <- carto_pal(12, "Safe")
ticker_colors <- c("#88CCEE", "#CC6677", "#DDCC77" ,"#117733", "#332288" ,"#AA4499","#999933" ,"#44AA99", "#882255", "#6699CC", "#888888")

```

```{r, eval=FALSE}
top50_symbol = c("AAPL", "NVDA", "MSFT", "AMZN", "META", "GOOGL", "BRK-B","AVGO", 
             "GOOG", "LLY", "JPM", "TSLA", "XOM", "UNH", "V", "MA", "HD", "PG", 
             "COST", "JNJ", "WMT", "NFLX", "ABBV", "CRM", "BAC", "ORCL", "MRK", 
             "KO", "CVX", "AMD", "PEP", "ACN", "LIN", "CSCO", "TMO", "MCD", 
             "ADBE", "WFC", "IBM", "GE", "ABT", "CAT", "NOW", "QCOM", "PM", 
             "ISRG", "VZ", "TXN", "DIS", "DHR")
```

```{r, eval=FALSE}
pull_stock_data <- function(symbol) {
  stock_data <- tryCatch({
    tq_get(top50_symbol, from = "2022-12-31") %>%
      as.data.frame() %>% 
      mutate(daily_return = (adjusted / lag(adjusted) - 1) * 100) %>% na.omit()
  }, error = function(e) {
    return(NULL)
  })
  
  
  return(stock_data)
}
```

```{r, eval=FALSE,warning=FALSE}
timer <- createTimer(precision = "ms") 
timer$start("e")

data <- pull_stock_data(top50_symbol)

timer$stop("e")

stock_data_list <- split(data, data$symbol)


write_xlsx(stock_data_list, "stock_data_by_company.xlsx")


```


```{r}
```{r}
# Load required libraries
library(ggmap)
library(dplyr)

# Set up Google Maps API key (replace 'your_api_key' with your actual API key)
register_google(key = "AIzaSyD3wR5ZmNZmv36w-LbmWixO5JKYmdWRtoI")

# Function to get latitude and longitude
get_lat_long <- function(address) {
  result <- geocode(address, output = "latlona", source = "google")
  return(result)
}

# Add a full address column to the 'current' data for geocoding
current <- current %>%
  mutate(full_address = paste(headquarters_location, "USA", sep = ", "))

# Fetch latitude and longitude for each company's headquarters
current <- current %>%
  rowwise() %>%
  mutate(geo_data = list(get_lat_long(full_address))) %>%
  mutate(latitude = geo_data$lat, longitude = geo_data$lon) %>%
  select(-geo_data)

# Print the updated dataset with latitude and longitude
print(head(current))

```


```{r}
# Load required libraries
library(ggmap)
library(dplyr)
current <- read.csv("data/company_data_full.csv")
head(data)
```


```{r}
# Set up Google Maps API key (replace 'your_api_key' with your actual API key)
register_google(key = "AIzaSyD3wR5ZmNZmv36w-LbmWixO5JKYmdWRtoI")

# Function to get latitude and longitude
get_lat_long <- function(address) {
  result <- geocode(address, output = "latlona", source = "google")
  return(result)
}

# Add a full address column to the 'current' data for geocoding
current <- current %>%
  mutate(full_address = paste(headquarters_location, "USA", sep = ", "))

# Fetch latitude and longitude for each company's headquarters
current <- current %>%
  rowwise() %>%
  mutate(geo_data = list(get_lat_long(full_address))) %>%
  mutate(latitude = geo_data$lat, longitude = geo_data$lon) %>%
  select(-geo_data)

# Print the updated dataset with latitude and longitude
print(head(current))

```


